\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-27 metatheory-condensed.tex
fix unicode chars in listing

We contribute \TT, a core type theory that incorporates the core elements
of \underline{f}amily polymorphism into
\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory~\cite{martin1982constructive},
while maintaining consistency and canonicity.

@@

As is standard in MLTT presentations, we use explicit substitutions~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are represented by de Bruijn indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{(\lam{\var{1}})}$.
%
Typing judgments for substitutions have the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$, which says that applying~$\gm$ to terms
valid in the context~$\Dl$ yields terms valid in~$\Gm$. 
%
The two main forms of substitutions are weakening and extension:
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/proj} states that if $t$ is a dependent pair
of type $\TySigma{A}{B}$, then $\snd{t}$ has type $\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$.

@@

    (* s₁  : {} | {} *)
    - o₁  = μ+ {} s₁ ⋆ : 𝕃{ty : 𝕎τ₁}
    (* s₂  : {ty : 𝕎τ₁} | [ty : 𝕌; bool : ty; ...] *) 
    - o₂  = μ+ o₁ s₂ ⋆ : 𝕃{ty : 𝕎τ₁; tm : 𝕎τ₂}
    (* s₃₁  : {ty : 𝕎τ₁; tm : 𝕎τ₂} 
            | [ty : 𝕌; ... ; tm : 𝕌; tvar : tm; ...] *)
    - o₃₁  = μ+ o₂ s₃ (μ+ ..) : 𝕃{ty : 𝕎τ₁; tm : 𝕎τ₂; subst' : 𝕃(RecSig)}
    (* s₃ : {ty : 𝕎(τ₁); tm : 𝕎τ₂; subst' : 𝕃(RecSig)} 
          | [ty : 𝕌; ... ; tm : 𝕎τ₂; subst' : 𝕃(RecSig)] *)
    - o₃ = μ+ o₃₁ s₃ (λ (Wrec ..)) : 𝕃{ty : 𝕎τ₁; tm : 𝕎τ₂; subst' : 𝕃(RecSig); subst : tm → nat → tm → tm}

    (* s₄ : {ty : 𝕎τ₁; ... subst : tm → nat → tm → tm } 
          | [ty : 𝕌; ... ; tm : 𝕌; subst : tm → nat → tm → tm] *)
    - o₄ = μ+ o₃ s₄ .. : 𝕃{ty : 𝕎τ₁; ... t₁ : tm}

    (* s₅ : {ty : 𝕎τ₁; tm : 𝕎τ₂; subst : 𝕃(RecSig); t1 : tm} 
          | [ty : 𝕌; tm : 𝕌; subst : 𝕃(RecSig); t1 : tm] *)
    - o₅ = μ+ o₄ s₅ : 𝕃{ ... }
  \end{lstlisting}

@@

Each hiding comment corresponds to the textual-aligned sealing derivation on RHS. For eligibility, curly brackets indiate a signature (omitting sealing at type level for readability) and square bracket indicate a nested sigma type, corresponding to sealing judgement $\goodSeal{}{s}{\sigma}{A}$.

>>>>

We contribute \TT, a core type theory that incorporates the core elements
of \underline{f}a\underline{m}i\underline{l}y polymorphism into
\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory~\cite{martin1982constructive},
while maintaining consistency and canonicity.

@@

As is standard in MLTT presentations, we use explicit substitutions~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are represented by de Bruijn indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{(\lam{\var{1}})}$.
%
Typing judgments for substitutions have the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (e.g., rule \ruleref{tm/sub}).
%
The two main forms of substitutions are weakening and extension:
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/proj} states that if $t$ is a dependent pair
of type $\TySigma{A}{B}$, then $\snd{t}$ has type $\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$.

@@

    (* s[@₁@]  : {} | {} *)
    - o[@₁@]  = μ+ {} s[@₁@] ⋆ : 𝕃{ty : 𝕎τ[@₁@]}
    (* s[@₂@]  : {ty : 𝕎τ[@₁@]} | [ty : 𝕌; bool : ty; ...] *) 
    - o[@₂@]  = μ+ o[@₁@] s[@₂@] ⋆ : 𝕃{ty : 𝕎τ[@₁@]; tm : 𝕎τ[@₂@]}
    (* s[@₃@][@₁@]  : {ty : 𝕎τ[@₁@]; tm : 𝕎τ[@₂@]} 
            | [ty : 𝕌; ... ; tm : 𝕌; tvar : tm; ...] *)
    - o[@₃@][@₁@]  = μ+ o[@₂@] s[@₃@] (μ+ ..) : 𝕃{ty : 𝕎τ[@₁@]; tm : 𝕎τ[@₂@]; subst' : 𝕃(RecSig)}
    (* s[@₃@] : {ty : 𝕎(τ[@₁@]); tm : 𝕎τ[@₂@]; subst' : 𝕃(RecSig)} 
          | [ty : 𝕌; ... ; tm : 𝕎τ[@₂@]; subst' : 𝕃(RecSig)] *)
    - o[@₃@] = μ+ o[@₃@][@₁@] s[@₃@] (λ (Wrec ..)) : 𝕃{ty : 𝕎τ[@₁@]; tm : 𝕎τ[@₂@]; subst' : 𝕃(RecSig); subst : tm → nat → tm → tm}

    (* s[@₄@] : {ty : 𝕎τ[@₁@]; ... subst : tm → nat → tm → tm } 
          | [ty : 𝕌; ... ; tm : 𝕌; subst : tm → nat → tm → tm] *)
    - o[@₄@] = μ+ o[@₃@] s[@₄@] .. : 𝕃{ty : 𝕎τ[@₁@]; ... t[@₁@] : tm}

    (* s[@₅@] : {ty : 𝕎τ[@₁@]; tm : 𝕎τ[@₂@]; subst : 𝕃(RecSig); t1 : tm} 
          | [ty : 𝕌; tm : 𝕌; subst : 𝕃(RecSig); t1 : tm] *)
    - o[@₅@] = μ+ o[@₄@] s[@₅@] : 𝕃{ ... }
  \end{lstlisting}

@@

Each hiding comment corresponds to the textual-aligned sealing derivation on RHS. For legibility, curly brackets indicate a signature (omitting sealing at type level for readability) and square bracket indicate a nested sigma type, corresponding to sealing judgment $\goodSeal{}{s}{\sigma}{A}$.
